---
title: "**Shiny**"
subtitle: "Introduction to web application development with Shiny"
author: "[STOR 390](https://idc9.github.io/stor390/)"
output:
  html_document:
    theme: cosmo
    toc: yes
    toc_float: yes
---



You can make interactive web applications easily and entirely in R with Shiny.
Shiny apps are a great way to share your analysis with others without them having to 
run any R themselves.


This lecture will cover (**list in progress**)

- first Shiny app
- beer to burpees app
- the basics of reactivity
- renderUI
- advanced: 
    - file uploading & downloading
    - pdf/html/word generation
    - database connection
    - modules


.


# **Hello Shiny!**

Running the code below in the R console will start up a simple Shiny app that
plots a histogram and lets the user choose the number of bins in the histogram 
using a slider.  The code that powers the app is included, and relevant parts 
of the code will be highlighted yellow as you interact with the app.


```{r, eval=FALSE}
if (!require('shiny')) install.packages("shiny")
runExample("01_hello")
```

# **Key components of a Shiny app**

A Shiny app is composed of two fundamental parts:

- **ui** - This codes the __user interface__ that controls the layout and appearance of the app.
- **server** - This codes the computations and plots that show up in the user interface.

## File structure

These two parts can either be included in one file called **app.R** or 
in two separate files called **ui.R** and **server.R**.  Both methods work
equivalently, and it's a matter of preference which one you choose and may 
depend on if your app is very short or long.

This is the bare minimum code (it does nothing!) if the app is on one **app.R** file:

```{r, eval=FALSE}
library(shiny)
ui <- fluidPage()
server <- function(input, output) {}
shinyApp(ui = ui, server = server)
```

And as you may suspect already, for two separate files, **ui.R** will contain

```{r, eval=FALSE}
library(shiny)
ui <- fluidPage()
```

and **server.R** has

```{r, eval=FALSE}
library(shiny)
server <- function(input, output) {}
```


# **Additional materials**

* [RStudio Webinar - How to start with Shiny – Part 1](https://www.rstudio.com/resources/webinars/how-to-start-with-shiny-part-1/)
* [RStudio Shiny Examples](https://github.com/rstudio/shiny-examples)
* [Show Me Shiny App Gallery](https://www.showmeshiny.com/)
* [Building Shiny apps - an interactive tutorial](http://deanattali.com/blog/building-shiny-apps-tutorial/)








