---
title: "Predictive modeling"
author: "STOR 390"
output: slidy_presentation
---


# Capital Bikeshare

![[]](http://www.freetoursbyfoot.com/wp-content/uploads/2014/04/article_2010_1012_capitalbikeshare.jpg)

# Load the data
```{r, warning=FALSE, message=F}
library(tidyverse)
hour <- read_csv('https://raw.githubusercontent.com/idc9/stor390/master/data/bikes_2011.csv')
```




# Look at the data

```{r}
head(hour)
```


# rider counts
```{r}
ggplot(hour) +
    geom_histogram(aes(x=cnt))
```


# temp vs. count
```{r}
ggplot(hour) +
    geom_point(aes(x=temp, y=cnt))
```


# riders per hour

```{r}
ggplot(hour) +
    geom_point(aes(x=hr, y=cnt))
```


# Consider a smaller problem

```{r}
hour <- hour %>% 
    select(cnt, hr)
```





# mean trend curve
```{r}
hour %>% 
    group_by(hr) %>% 
    summarise(mean_cnt=mean(cnt)) %>% 
    ggplot() +
    geom_line(aes(x=hr, y=mean_cnt)) +
    geom_point(aes(x=hr, y=mean_cnt))
```




# Predictive modeling

- predict the number of riders based only on the time of day 

    - build a model mapping `hr` to `cnt`

# Linear regression
You should always start with linear regression!
```{r}
ggplot(hour) +
    geom_point(aes(x=hr, y=cnt)) +
    geom_smooth(aes(x=hr, y=cnt), color='red', method=lm, se=FALSE)
```


# Let's build more models to do better
- non linear models
- many possibilities


# Decide on a metric
- evaluate each model with one number to decide how it's doing
- recall *residuals* from the last lecture
- mean squared error (MSE)



# MSE for linear model

```{r}
linear_model <- lm(cnt ~ hr, hour)

# put the actual and predicted counts in a data frame
results <- tibble(cnt_actual = hour$cnt,
                  cnt_pred=linear_model$fitted.values)

results %>% 
    mutate(resid=cnt_actual - cnt_pred) %>% 
    mutate(resid_sq = resid^2) %>% 
    summarise(MSE=mean(resid_sq))

```


# Overfitting

> a statistical model describes random error or noise instead of the underlying relationship. 

(from wikipedia)

# is bad...

> A model that has been overfit has poor predictive performance, as it overreacts to minor fluctuations in the training data.


# French class

- Je n'aime pas la classe française

# Overfitting movies

```{r, echo=F, warning=F, message=F}
movies <- read_csv('https://raw.githubusercontent.com/idc9/stor390/master/data/movies.csv')

movies[movies[, 'title' ] == 'The End of America', 'runtime'] <- 73

data <- movies %>% 
        select(imdb_rating, imdb_num_votes,
               critics_score, audience_score,
               runtime, genre, mpaa_rating,
               best_pic_win) %>% 
        mutate(genre=factor(genre),
               mpaa_rating=factor(mpaa_rating), 
               best_pic_win=factor(best_pic_win))

data_trans <- data %>% 
                mutate(nv_sqrt = sqrt(imdb_num_votes),
                       nv_sq = imdb_num_votes^2,
                       nv_cube = imdb_num_votes^3,
                       nv_log = log(imdb_num_votes))  
                # select(imdb_rating, imdb_num_votes, nv_sqrt, nv_sq, nv_cube, nv_log)


lin_reg_trans <- lm(imdb_rating ~., data_trans)


pred_df <- tibble(imdb_rating_pred = unname(predict(lin_reg_trans)),
                  imdb_num_votes=data_trans$imdb_num_votes,
                  imdb_rating=data_trans$imdb_rating)

ggplot(pred_df) +
    geom_point(aes(x=imdb_num_votes, y=imdb_rating)) +
    geom_line(aes(x=imdb_num_votes, y=imdb_rating_pred), color='red')

```



# Two distinct concepts

- model building
- model evaluation



# Exploratory vs confirmatory

> Each observation can either be used for exploration or confirmation. 

# Model building vs. model evaluation

> Each observation can either be used for building a model or evaluating a model, not both.



# Train vs. test set

```{r}
# there are n observations
n <- dim(hour)[1]

# number of observations that go in the training st
n_tr <- floor(n * .8)


# randomly select n_tr numbers, without replacement, from 1...n
tr_indices <- sample(x=1:n, size=n_tr, replace=FALSE)

# break the data into a non-overlapping train and test set
train <- hour[tr_indices, ]
test <- train[-tr_indices, ]
```

# Fit a bunch of models

```{r}
# manually add hr^2 to the data matrix
train_square <- mutate(train, hr_sq = hr^2)
model_square <- lm(cnt ~ hr + hr_sq, train_square)

# there is a better way to do this using R's modeling language
model_square <- lm(cnt ~ hr + I(hr^2), train)
```

# Let's do this automatically

```{r}
# largest degree polynomial to try
d_max <- 20

# lets save each model we fit in a list
models <- list()

# also store the traing error in data frame
error <- tibble(degree=1:d_max,
                MSE_tr = rep(0, d_max))

# fit all the models
for(d in 1:d_max){
    # the poly function does exactly what you think it does
    models[[d]] <- lm(cnt ~ poly(hr, d), train)
    
    # compute the MSE for the training data
    mse_tr <- mean(models[[d]]$residuals^2)
    
    # save the MSE
    error[d, 'MSE_tr'] <- mse_tr
}

error
```








# Training error
```{r}
# plot the training error
ggplot(error)+
    geom_point(aes(x=degree, y=MSE_tr)) +
    geom_line(aes(x=degree, y=MSE_tr))

```


# Degree 20 model
```{r, echo=F}
model_20 <- models[[20]]

# get the predictions at each hour
pred <- tibble(hr=0:23)
hr_predictions <- predict(model_20, newdata = pred)              


pred <- pred %>% 
        mutate(cnt_pred =hr_predictions)

# plot the predictions over the training data
ggplot(data=train)+
    geom_point(aes(x=hr, y=cnt)) +
    geom_line(data=pred, aes(x=hr, y=cnt_pred), color='red')

```



# Test error

```{r, echo=F}
# lets add the test error to the error data frame
error <- error %>% 
    add_column(MSE_tst=rep(0, d_max))


for(d in 1:d_max){
    
    # grab the trained model
    model <- models[[d]]
    
    # get the predictions for the test data, compute the residuals
    
    test_results <- test %>% 
           mutate(hr_pred = predict(model, newdata=test)) %>% 
           mutate(resid_sq = (hr-hr_pred)^2) 

    # compute the MSE
    mst_tst <- summarise(test_results, mse_tst = mean(resid_sq))[[1]]

    error[d, 'MSE_tst'] <- mst_tst
}


ggplot(error)+
    geom_point(aes(x=degree, y=MSE_tst)) +
    geom_line(aes(x=degree, y=MSE_tst)) 

```



# Train vs test

```{r}
error %>% 
    rename(tr=MSE_tr, tst=MSE_tst) %>% 
    gather(key=type, value=error, tr, tst) %>% 
    ggplot() +
    geom_point(aes(x=degree, y=log10(error), color=type)) +
    geom_line(aes(x=degree, y=log10(error), color=type))

```













