---
title: "Harry Potter and regular expressions"
author: 'STOR 390'
output: html_document
---

Assignment 2 is due 2/28/17. You can find the raw .Rmd file at: [https://raw.githubusercontent.com/idc9/stor390/master/assignments/harry_potter/harry_potter.Rmd](https://raw.githubusercontent.com/idc9/stor390/master/assignments/harry_potter/harry_potter.Rmd).

The text of all 7 Harry Potter books is available online: [http://www.readfreeonline.net/Author/J._K._Rowling/Index.html](http://www.readfreeonline.net/Author/J._K._Rowling/Index.html). In this assignment you will use dplyr, ggplot and regular expressions to do an exploratory analysis of Harry Potter and the Philosopher's Stone.

Here are a couple examples of similar text analysis projects (that you will be able to do in a couple weeks!)

- [The Life-Changing Magic of Tidying Text](http://juliasilge.com/blog/Life-Changing-Magic/) by Julia Silge (yes [janeaustenr](https://github.com/juliasilge/janeaustenr) is an entire R package devoted to Jane Austen)

- [Harry Potter agression](https://github.com/andrewheiss/Harry-Potter-aggression) by Andrew Heiss

# Load the data
The `rvest` package makes scraping websites easy with R. The following code will download the text of the Harry Potter and the Philosopher's Stone from readfreeonline and save it as a string (called `text`). I suggest running this code once then saving text on your computer and using `read_lines` to read the text file into your computer.

```{r, eval=F}
library(tidyverse)
library(rvest)


chapter_base_url <- 'http://www.readfreeonline.net/OnlineBooks/Harry_Potter_and_the_Sorcerers_Stone/Harry_Potter_and_the_Sorcerers_Stone_'
num_chapters <- 17

# download each chapter and concatonate them
text <- ''
for(i in 1:num_chapters){
    
    # url for the chapter text
    chapter_url <- paste0(chapter_base_url, i, '.html')
    
    # download the chatpter html
    chapter_html <- read_html(chapter_url)  
    
    # extract the text of the chatper from the html
    chapter_text <- chapter_html %>%
                html_nodes(".ContentCss") %>%
                html_text()
    
    #
    text <- paste0(text, chapter_text)
}

# save as a .txt file
# write_lines(text, 'philosophers_stone.txt')
```


# Question 0
Set `eval=FALSE` for the chunck above and `eval=TRUE`for the chunck below and all test chunks.
```{r, eval=F, message=F, warning=F}
# set up
library(tidyverse)
library(stringr)
text <- read_file('philosophers_stone.txt')
```

# Question 1
How many words are in the book? 

```{r}
#
```


# Question 2 

How many times are each of the following characters mentioned? Display the answer using an appropriate visualization. 

    - Harry, Hermione, Ron, Neville, Dumbledore, Draco, Snape, Hagrid, McGonagall


*Hint*: the `map` function might be helpful (but not necessary).


```{r}
people <- c('Harry', 'Hermione', 'Ron', 'Neville', 'Dumbledore', 'Draco', 'Snape', 'Hagrid', 'McGonagall')

```






# Question 3
Break the text into paragraphs; create a verctor called `paragraphs` where each entry is a paragraph in the book.


```{r}
#
```


# Question 4
Write a function that can break the text up into paragraphs, sentences, or words. This is a preview of [what you'll be doing](http://tidytextmining.com/tidytext.html#the-unnest_tokens-function) in a couple weeks.

This function does not need to be perfect. For sentences, give one example where the function you wrote fails.

*Hint*: the function should probably have a if statement
```{r}
unnest_tokens <- function(text, token='words'){
    # splits a string into tokens
    # input
        # text is a string
        # token can be one of: words, paragraphs, sentences
    # output: a character vector
    
}

```


```{r test4, eval=F}
# TODO add more
# Test code for the grader -- you don't have to modify these
sum(paragraphs == unnest_tokens(text, 'paragraphs'))
```


# Question 5

Put the data into tidy format with one row per paragraph. 

- first remove all paragraphs length 0
- create a tibble called `paragraph_df` with one column `text` with the text of each paragraph (*hint*: you might need to use `as.character(paragraphs)`)
- add a new column `index` that gives the index of each paragraph
- **wighout** using dplyr add a column called `Harry` that counts the number of times Harry is referenced in each paragraph


```{r}
#
```

*Hint*: you can use question 2 to check your answer

# Question 6
Write a function called `reference_counter` that generalizes question 5 for any tidy text data frame and any list of words. 

*Hint*: do this **without** dplyr

```{r}

reference_counter <- function(text_df, word_list){
    
    # inputs
        # text_df is a tibble with a column called text
        # word_list is a vector of strings
    # for each word in word_list add a column to text_df counting
    # the number of times that word appears in each row of text df
    # does not modify the original text_df
    # do this WITHOUT using dplyr

}
```


```{r test6, eval=F}
# test code for grader
test_words <- c('Harry', 'Hagrid', 'wand')
test_df <- reference_counter(paragraph_df, test_words)

test_df %>% select(Harry, Hagrid, wand) %>% summarise_all(sum)
```



# Question 7 

Using the `reference_counter` function update `paragraph_df` to include columns counting the number of references to each characters from Q2 in each paragraph


```{r}
#
```



```{r test7, eval=F}
# test code for grader
paragraph_df[,people] %>% summarise_all(sum)
```


# Question 8

Make a new data frame called `person_refs`  with three columns: person, num_refs, index. num_refs is the number of references each person gets in paragraph and index is the index of the paragraph. Limit this data frame to the following 5 characteres: Harry, Hermione, Ron, Draco, Neville. 

*Hint*: use `gather`.

```{r}
#
```


Make a bar plot showing the number of paragraphs that references each of the 5 characters

```{r}
#
```


Now we want to examine how characters evolove over "time." Plot the number of references vs. the paragraph index.

```{r}
#
```

In this question we are using paragraphs for "time windows." What are other "time windows" we could have used? What are some tradeoffs for these different choices.

# Question 9
How often are Harry and Herminone referenced together? Plot the number of references per paragraph for Harry vs. Herminone.

- one plot using `geom_point`
- one plot using `geom_jitter` (usethe width/height arguments of jitter to make the jitter plot look better)


```{r}
#
```


```{r}
#
```


Why is the jitter plot better than a simple point plot?


# Question 10
Do Harry and Hermione tend to co-occur?  Fit a linear regression of Harry vs. Hermione references per paragraph. Use the `lm()` function and print out the `summary` of the model.


```{r}
#
```


Now use `geom_smooth` to plot the linear regression line on top of the jitter plot.

```{r}
#
```


# Question 11
Is there are relationship between the length of the pargraph a the number of times Harry is mentioned? Add a column called `num_words` to `paragraph_df` counting the number of words in each paragraph. Then use a linear regression to answer for the question. Provide both a statistical summary and a visualization.


```{r}
#
```

# Question 12
Create an indeicator variable `harry_mentioned` that indicates whether or not Harry is mentioned in each paragraph. This indicator variable should be a factor (e.g. use `as.factor`).

```{r}
#
```

Now repeate the previous linear regression with `harry_mentioned` as the x variable instead of the number of times he is mentioned

```{r}
#
```


# Free response
Ask and answer a question with this data set. You should make at least 2 figures (e.g. plot, printout of a regression, etc). Provide a written explaination of the question and the evidence for your answer.









