---
title: "Assignment 1: dplyr and UNC departments"
author: "STOR 390"
output: html_document
---
Due **Tuesday 1/31/17** before the start of class.

Universities are typically opaque, bureaucratic institutions. Many public schools such as the University of North Carolina system make [data about university employees publically](http://www.newsobserver.com/news/databases/public-salaries/) available. Using this dataset what can we learn about the dynamics of the university system?


## dplyr

To download the .Rmd file for this assignment go to the [github repo](https://github.com/idc9/stor390/tree/master/assignments/unc_depts), right click **und_depts.Rmd** and select *save link as*.

The best way to to learn the dplyr package is to just use it a lot. Read through the following dplyr tutorial the complete questions in this .Rmd document. I suggest doing the coding in an R script then putting the answers into this document.

- [dplyr flights vignettes](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html)

- chapter 5 ([data transformation](http://r4ds.had.co.nz/transform.html)) from r4ds

## Instructions
- please finish these questions by Tuesday, Jan 24. More questions will added then.
- use the `dplyr` functions to answer the following questions (base R will not be given credit) 
- submit just the .Rmd file to Sakai
- pay careful attention to the questions

- **return** something means just print it out. Similarly for questions **What is ...** just print out the answer

Q: Return the sum of the first 10 positive integers?

Q: What is the sum of the first 10 positive integers?
```{r}
sum(1:10)
```

- **create** or **save** means save a variable

Create a variable `x` with the sum of the first 10 positive integers.
```{r}
x <- sum(1:10)
```



# The data

First download the data set from Ryan Thornburg's website (it is also on stor390's github repo). You may want to set `cache=TRUE` for this chunk so you don't have to download the data set every time you run the script.
```{r setup, message=FALSE, cache=FALSE}
library(tidyverse)

# Load the UNC departments data
data <- read_csv(url("http://ryanthornburg.com/wp-content/uploads/2015/05/UNC_Salares_NandO_2015-05-06.csv"))
```

Take a first look at the data
```{r}
colnames(data)

str(data)

head(data)
```

# hints

- Read through the first dplyr tutorial. Every question can be answered using: filter, arrange, select, mutate, summarize and group_by functions.

- here are two more dplyr tutorials: https://rpubs.com/justmarkham/dplyr-tutorial and https://www.coursera.org/learn/data-cleaning/home/week/3

- fte means full time employee


# Assignment


## Question 1
Return a data frame with columns: name, dept, age,totalsal 
```{r}
#
```

Return a data frame with columns: statesal, nonstsal,totalsal using one of the `select_helpers` functions
```{r}
#
```

## Question 2 
Rename the `fte` column to `fulltime`. Make sure this change is saved (i.e. `data <- ...`).
```{r}
#
```

## Question 3 

What is the mean salary in the Neurosurgery department?

```{r}
#
```


Return a data frame with employee's in the Neurosurgery department making more than $500,000. Why might these professors be so well paid? 
```{r}
#
```



## Question 4 
What is the total amount that full time Dermatology employees get paid?


```{r}
#
```


## Question 5

Create a data frame called radio_dept whose rows are the employees from the Radiology department. 

- include only the following columns: name, position, age, nonstsal, totalsal. 
- order the employees by salary

First without chaining

```{r }
#
```


Next with chaining
```{r }
#
```


Make a histogram of Radiology salaries
```{r }
#
```


## Question 6

Use faceting to make totalsal histograms for the Radiation Oncology, Radiology, Pediatrics, and Orthopaedics departments.
```{r }
#
```



Use faceting to make plot age vs. totalsal for the same four departments and color points by status.
```{r }
#
```

## Question 7

Create a data frame called `dept_summary`  whose rows are the departments and  whose columns are: department size, mean department salary, median department salary, and maximum salary (using totalsal for salary). 

```{r }
#
```

Order the departments by highest mean salary and print the 10 highest paid departments. 


```{r }
#
```

Order the departments by highest median salary and print the 10 highest paid departments. 

```{r }
#
```

Why do these lists differ? If you were asked for the top 10 best paid departments at UNC which summary would you choose and why?

## Question 8
How many departments have at least 10 employees?
```{r }
#
```


## Question 9
Which department hired the most people in 2010? Hint: you probably have to modify `hiredate`.
```{r }
#
```

## Question 10
Make a list of all the department names and sort this list alphabetically. What is the 42nd department in this list?
```{r }
#
```


## Question 11
Plot number of people hired by the CS dept per year vs. year
```{r }
#
```


Now add STOR, Math, Biostatistics, SILS and Physics to the above plot
```{r }
#
```


# Open question

Examine the relationship between age and salary at UNC. You can interpret this however you like e.g.

- age can mean number of years employed at the university
- compare and contrast different departments

Requirements

- hand in a separate .Rmd document 
    - load the data with read_csv(url(...)) like this script
- make 3 figures
    - a figure can be a visualization, print out of a data frame that you can created, the results of a statistical model/test (e.g. linear regression)
- display all code used
    - code should be commented
    - if you have a lot of functions you can put them in a separate .R script
- describe each figure
    - what is the figure (figure blah is a plot of foo vs. bar)
    - what is the upshot of the figure (figure blah shows that there is a strong relationship between foo and bar)
- write an explanation of your findings
    - a couple paragraphs
    - use proper grammar
    - your findings should be clear
