---
title: "Lab : Classification"
output: html_document
---

This lab will give you practice with

- cross-validation
- test/train set
- K nearest neighbors
- nearest centroid classifier

# KNN cross-validation


Using cross-validation we found that K = 1 is the "best" value for KNN on the Human Activity Recogntion dataset. But is this truly the best value of K?


Let's call the plot of error vs. K the *tuning error curve*.

## Test error


1. Compute the test set error (using the independent test data) for each value of K. 
2. Plot the test set error with the the cross-validation test set tuning error curve (from lecture).
3. What is actually the best value of K?
4. How well does the cross-validation error approximate the true test set error?

## What happens when we change the number of folds

1. Compute the training error for each value of K.
2. Add the training error to the tuning error curves (there should now be three tuning error curves).
3. Write a function that makes this tuning error curve plot.
4. Make the tuning error plots for 5, 10, 20, and 50 folds.

```{r}
knn_tuning_error_plot <- function(train, test, k_cv, k_values, cv_seed=NA){
# Returns the tuning error plot for KNN
# train and test are the train and test data
    # both are a data frame with the same column names
    # one column in named y which is the class labels
# k_cv is the number of cross validation folds
# k_values is the sequence of K values try for KNN
# cv_seed is the seed for the cross validation folds
# returns a ggplot object    
    
    # set seed if it is given
    if(!is.na(cv_seed)){
        set.seed(cv_seed)
    }

    
    
    # p <- ggplot() + ...
    # return(p)

}

```


# Nearest Centroid



## Nearest centroid function
Write a function that implements the nearest centroid classifier. The function should take the training and test data as input and return the predictions on the test data set.




```{r}
nearest_centroid <- function(train_x, train_y, test_x){
    # returns the predictions for nearest centroid on a test set
    # train_x and test_x are the train/test x data
        # assume these are both numerical matrices with the same number of columns
    # train_y is a vector of class labels for the training data
    # return a vector of predicted class labels for the test data
}


```


## Test error

Using the `nearest_centroid` function compute the test set error for the Human Activity Recognition dataset.






