---
title: "Lab 5: joins"
output: html_document
---
The lab is due Thursday 2/9/17. You can download this .Rmd file from []()

# Read in some toy joins data

```{r, message=F, warning=F}
library(tidyverse)
banking_account_types <- read_csv('https://raw.githubusercontent.com/idc9/stor390/master/data/joins/banking_account_types.csv')
banking_transactions <- read_csv('https://raw.githubusercontent.com/idc9/stor390/master/data/joins/banking_transactions.csv')


crime_arrests <- read_csv('https://raw.githubusercontent.com/idc9/stor390/master/data/joins/crime_arrests.csv')
crime_wealth <- read_csv('https://raw.githubusercontent.com/idc9/stor390/master/data/joins/crime_wealth.csv')

titanic_outcomes <- read_csv('https://raw.githubusercontent.com/idc9/stor390/master/data/joins/titanic_outcomes.csv')
titanic_person_features <- read_csv('https://raw.githubusercontent.com/idc9/stor390/master/data/joins/titanic_person_features.csv')

toy1 <- read_csv('https://raw.githubusercontent.com/idc9/stor390/master/data/joins/toy_one.csv')
toy2 <- read_csv('https://raw.githubusercontent.com/idc9/stor390/master/data/joins/toy_two.csv')

play1 <- read_csv('https://raw.githubusercontent.com/idc9/stor390/master/data/joins/play_one.csv')
play2 <- read_csv('https://raw.githubusercontent.com/idc9/stor390/master/data/joins/play_two.csv')
```

For each of the following questions do the specified joins. Remember to follow the process looking at header, data set row counts and data types in advance and make a prediction about the join outcome. Print out the dimensions of the resulting data frames.

# Question 1
Join the `toy` data sets using inner, outer, and left joins.  

```{r}
#
```

```{r}
#
```

```{r}
#
```

# Question 2
Join the `play` data sets using inner, outer, and left joins.


```{r}
#
```

```{r}
#
```


```{r}
#
```

# Question 3
Join the `banking` data sets using inner join.


```{r}
#
```

# Question 4
Join the `crime` data sets using inner join.


```{r}
#
```

# Question 5
Join the `titanic` data sets using inner join.


```{r}
#
```

