---
title: "STOR 320 Shiny"
author: | 
  | Frances Tong
  | Data Scientist
  | BD Technologies & Innovation
  | 
  | Talk URL
  | [https://francestong.shinyapps.io/stor320_shiny/](https://francestong.shinyapps.io/stor320_shiny/)
  | Code of talk 
  | [https://github.com/francestong/stor390/tree/master/stor320_shiny/stor320_shiny.Rmd](https://github.com/francestong/stor390/tree/master/stor320_shiny/stor320_shiny.Rmd)
runtime: shiny
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Who am I?

* PhD in statistics
* Analysis of genomic data 
* [list kinds of projects]
* Demo of company app

## Topics today

[list topics]
[include link to jump there]

## renderUI 

One of the more powerful render functions is renderUI() - it enables us to 
generate dynamic UI, i.e. construct reactive objects that depend on user input. 
We illustrate how it works with a toy example that requires this function.

Let's say we are selecting colors for plotting in R, and we want to help the 
user narrow down the choices through color families.

If the user wants a red color, then the drop-down menu can display only the
red options.  We can't put a `selectInput` drop down menu in **ui** when we 
can't populate the choices argument with the correct list of colors that 
depend on what the user chooses in the `radioButtons`.  This means that we need 
to dynamically generate this `selectInput` inside `renderUI` (on the **server** side!)
based on the user-chosen color family `input$color_family`.

## renderUI example

```{r}
color_list <- list(red = c("tomato", "violetred", "firebrick"),
                   green = c("chartreuse", "forestgreen", "olivedrab"),
                   blue = c("navy", "dodgerblue", "torquoise"))

inputPanel(
  radioButtons(inputId = "color_family",
               label = "Choose a color family",
               choices = c("red", "green", "blue"))
)

renderUI({
  selectInput(inputId = "color",
              label = "Select color",
              choices = color_list[input$color_family])
})

renderText({
  paste("You chose this color:", input$color)
})
```


## renderUI example code

```{r, echo=TRUE, eval=FALSE}
library(shiny)

color_list <- list(red = c("tomato", "violetred", "firebrick"),
                   green = c("chartreuse", "forestgreen", "olivedrab"),
                   blue = c("navy", "dodgerblue", "torquoise"))

ui <- fluidPage(
    
    radioButtons(inputId = "color_family",
                 label = "Choose a color family",
                 choices = c("red", "green", "blue")),
    
    uiOutput(outputId = "color_selector"),
    
    verbatimTextOutput(outputId = "color_chosen")
    
)

server <- function(input, output) {
    
    output$color_selector <- renderUI({
        
        selectInput(inputId = "color",
                    label = "Select color",
                    choices = color_list[input$color_family])
        
    })
        
    output$color_chosen <- renderText({
        paste("You chose this color:", input$color)
    })
    
    
}

shinyApp(ui = ui, server = server)
```



## Reactive expressions

Reactive expressions are a way for your app to update efficiently when users
interact with it.  They are a bit nuanced to use - here's the RStudio tutorial on them https://shiny.rstudio.com/tutorial/lesson6/. 

In general, any code you put in a `render` function will continually be updated, 
even if you don't necessarily need them to be rerun every single time.  This will
slow down your app considerably.  If you want to use data inside a `render` function
that does not need to be updated every time if `input` doesn't change, 
then you can wrap your data inside a reactive expression that can be called inside `render`.
The reactive expression will only recalculate if `input` has changed.


- Reactive expressions depend on the status of `input` values or values from other reactive expressions.
- You can create reactive expressions with `reactive({})`.
- You can access a reactive expression with its name followed by `()`.
- You can only call a reactive expression within a `reactive` or `render` function.

## Reactive expression example

```{r, eval=TRUE, echo=FALSE}
inputPanel(
  
  sliderInput(inputId = "rate",
              label = "Choose rate",
              min = 0,
              max = 1, 
              value = 0.5),
  
  textOutput(outputId = "pct_std"),
  
  textOutput(outputId = "pct_half")
)

my_percent <- reactive({
  input$rate * 100
})

my_half_percent <- reactive({
  my_percent() / 2
})

renderText({
  paste0("This is my percent: ", my_percent(), "%")
})

renderText({
  paste0("This is my percent by half: ", my_half_percent(), "%")
})
```


## Reactive expression example code

```{r, eval=FALSE, echo=TRUE}
library(shiny)

ui <- fluidPage(
    
    sliderInput(inputId = "rate",
                label = "Choose rate",
                min = 0,
                max = 1, 
                value = 0.5),
    
    textOutput(outputId = "pct_std"),
    
    textOutput(outputId = "pct_half")
    
)

server <- function(input, output) {
    
    my_percent <- reactive({
        input$rate * 100
    })
    
    my_half_percent <- reactive({
        my_percent() / 2
    })
    
    output$pct_std <- renderText({
        paste0("This is my percent: ", my_percent(), "%")
    })
    
    output$pct_half <- renderText({
        paste0("This is my percent by half: ", my_half_percent(), "%")
    })
    
}

shinyApp(ui = ui, server = server)
```


## Leaflet example

```{r}
library(leaflet)


inputPanel(
  # numericInput("lat", "Latitude", value = 35.891681),
  # numericInput("lng", "Longitude", value = -78.8608968),
  # textInput("label", "Place", value = "BD Technologies & Innovation")
  numericInput("lat", "Latitude", value = 35.910274),
  numericInput("lng", "Longitude", value = -79.051004),
  textInput("label", "Place", value = "You are here!")
)

renderLeaflet({
  leaflet() %>%
    addTiles() %>%  # Add default OpenStreetMap map tiles
    addMarkers(lng = input$lng, lat = input$lat, popup = input$label)
})
```


## Leaflet example code

```{r, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
library(shiny)
library(leaflet)

ui <- fluidPage(
  numericInput("lat", "Latitude", value = 35.910274),
  numericInput("lng", "Longitude", value = -79.051004),
  textInput("label", "Place", value = "You are here!")
  leafletOutput("mymap")
)

server <- function(input, output) {
  output$mymap <- renderLeaflet({
    leaflet() %>%
      addTiles() %>%  # Add default OpenStreetMap map tiles
      addMarkers(lng = input$lng, lat = input$lat, popup = input$label)
  })
}

shinyApp(ui, server)
```



